plyranges: A grammar of genomic data transformation

### 2018-06-15 paper.tex
changes for GB sub

@@

\author{\Large Stuart Lee\vspace{0.05in} \newline\normalsize\emph{Department of Econometrics and Business Statistics, Monash University\\
\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
Medicine Division, Walter and Eliza Hall Institute}   \and \Large Di Cook\vspace{0.05in} \newline\normalsize\emph{Department of Econometrics and Business Statistics, Monash University}   \and \Large Michael Lawrence\vspace{0.05in} \newline\normalsize\emph{Bioinformatics and Computational Biology, Genentech, Inc.}  }

@@

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Stuart Lee (Department of Econometrics and Business Statistics, Monash University\\
\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
Medicine Division, Walter and Eliza Hall Institute) and Di Cook (Department of Econometrics and Business Statistics, Monash University) and Michael Lawrence (Bioinformatics and Computational Biology, Genentech, Inc.)},
             pdfkeywords = {bioconductor, grammar, genomes, data analysis},  
            pdftitle={plyranges: A grammar of genomic data transformation},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
\textbf{\authorfont Stuart Lee} \hskip 15pt \emph{\small Department of Econometrics and Business Statistics, Monash University\\
\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}\hspace*{0.333em}Molecular
Medicine Division, Walter and Eliza Hall Institute}   \par \textbf{\authorfont Di Cook} \hskip 15pt \emph{\small Department of Econometrics and Business Statistics, Monash University}   \par \textbf{\authorfont Michael Lawrence} \hskip 15pt \emph{\small Bioinformatics and Computational Biology, Genentech, Inc.}

@@

\vskip 8.5pt \noindent \emph{Keywords}: bioconductor, grammar, genomes, data analysis \par

@@

\noindent  \hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{genomic-relational-algebra}{%
\section{Genomic relational algebra}\label{genomic-relational-algebra}}

\hypertarget{data-model}{%
\subsection{Data model}\label{data-model}}

@@

The \texttt{plyranges} DSL is built on the core Bioconductor data
structure GRanges, which is essentially a constrained table, with fixed
columns for the chromosome, start and end coordinates, and the strand,
along with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment (figure
\ref{fig:GRanges}). GRanges balances flexibility with formal
constraints, so that it is applicable to virtually any genomic workflow,
while also being semantically rich enough to support high-level
operations on genomic ranges. As a core data structure, GRanges enables
interoperability between plyranges and the rest of Bioconductor.
Adhering to a single data structure simplifies the API and makes it
easier to learn and understand, in part because operations become
endomorphic, i.e., they return the same type as their input.

GRanges follows the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable describing the
observations. GRanges specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
dplyr. Thus, we define our algebra as an extension of the dplyr algebra,
and borrow its syntax conventions and design principles.

@@

\hypertarget{algebraic-operations}{%
\subsection{Algebraic operations}\label{algebraic-operations}}

@@

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including \emph{intersect}, \emph{union},
\emph{difference}, and \emph{complement}. There are two sets of these:
parallel and merging. The parallel operations map to infix operators,
which we surround with \emph{\%} symbols, a convention borrowed from R
syntax. For example, the parallel intersection (\emph{x \%intersect\%
y}) finds the intersecting range between \emph{xi} and \emph{yi} for
\emph{i} in \emph{1\ldots{}n}, where \emph{n} is the length of both
\emph{x} and \emph{y}. In contrast, the merging intersection
(\emph{intersect\_ranges(x, y)}) returns a new set of disjoint ranges
representing wherever there was overlap between a range in \emph{x} and
a range in \emph{y}. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a \emph{span} operator that takes the union
while filling any gap. The \emph{complement} operation is unique in that
it is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the \emph{between} parallel operation,
which finds the gap separating \emph{xi} and \emph{yi}. The binary
operations are callable from within arithmetic, restriction and
aggregation expressions.

@@

\hypertarget{design-principles}{%
\subsection{Design principles}\label{design-principles}}

\hypertarget{cognitive-consistency-and-fluency}{%
\subsubsection{Cognitive consistency and
fluency}\label{cognitive-consistency-and-fluency}}

We aim for our interfaces to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the user
can express workflows in the language of genomics. This motivates the
adoption of the tidy GRanges object as our central data structure. The
basic data.frame and \texttt{dplyr} tibble lack any notion of genomic
ranges and so could not easily support our genomic grammar, with its
specific verbs for range-oriented data manipulation. Another example of
cognitive consistency is how \texttt{plyranges} is insensitive to
direction/strand by default when, e.g., detecting overlaps.
\texttt{GenomicRanges} has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

\hypertarget{cohesion}{%
\subsubsection{Cohesion}\label{cohesion}}

A function is cohesive if it performs a singular task without producing
any side-effects. Singular tasks are not necessarily atomic; they can
always be broken down further at lower levels of abstraction. For
example, to resize a range, the user needs to specify which position
(start, end, midpoint) should be invariant over the transformation. The
\texttt{resize()} function from the \texttt{GenomicRanges} package has a
\texttt{fix} argument that sets the anchor, so calling \texttt{resize()}
coalesces anchoring and width modification. The coupling at the function
call level is justified since the effect of setting the width depends on
the anchor. However, plyranges increases cohesion and decouples the
anchoring into its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names as
the primary means of expression, instead of a more complex mixture of
function and argument names. Since functions are superordinate to their
arguments, flattening the API at the function level enables the user to
conceptualize the API as a catalog of functions, without having to
descend further. A flat function catalog also enhances API
discoverability, particularly through auto-completion in IDEs. One
downside of pushing cohesion to this extreme is that function calls
become coupled, and care is necessary to treat them as a group when
modifying code.

\hypertarget{expressiveness}{%
\subsubsection{Expressiveness}\label{expressiveness}}

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect in
the original \texttt{GenomicRanges} API. In other cases, the
\texttt{plyranges} API increases verbosity for the sake of clarity and
cohesion. Explicitly calling \texttt{anchor()} can require more typing,
but the code is easier to comprehend. Another example is the set of
routines for importing genomic annotations, including
\texttt{read\textbackslash{}\_gff()},
\texttt{read\textbackslash{}\_bed()}, and \texttt{read\_bam()}. Compared
to the generic \texttt{import()} in \texttt{rtracklayer}, the explicit
format-based naming in plyranges clarifies intent and the type of data
being returned. Similarly, every plyranges function that computes with
strand information indicates its intentions by including suffixes such
as \texttt{directed}, \texttt{upstream} or \texttt{downstream} in its
name, otherwise strand is ignored. The \texttt{GenomicRanges} API does
not make this distinction explicit in its function naming, instead
relying on a parameter that defaults to strand sensitivity, an arguably
unintuitive behavior.

@@

\hypertarget{peak-finding}{%
\subsection{Peak Finding}\label{peak-finding}}

@@

\hypertarget{computing-windowed-statistics}{%
\subsection{Computing Windowed
Statistics}\label{computing-windowed-statistics}}

@@

\hypertarget{quality-control-metrics}{%
\subsection{Quality Control Metrics}\label{quality-control-metrics}}

@@

\hypertarget{discussion-and-conclusion}{%
\section{Discussion and conclusion}\label{discussion-and-conclusion}}

@@

A caveat to constructing a compatible interface with \texttt{dplyr} is
that \texttt{plyranges} makes extensive use of non-standard evaluation
in R via the \texttt{rlang} package \cite{R-rlang}. Simply, this means
that computations are evaluated in the context of the GRanges objects.
Both dplyr and plyranges are based on the \texttt{rlang} language,
because it allows for more expressive code that is free of repeated
references to the container. Implicitly referencing the container is
particularly convenient when programming interactively. Consequently,
when programming with \texttt{plyranges}, a user needs to generally
understand the \texttt{rlang} language and how to adapt their code
accordingly. Users familiar with the tidyverse should already have such
knowledge.

@@

As the \texttt{plyranges} interface encourages tidy data practices, it
integrates well with the grammar of graphics \cite{Wickham2016-gz}. To
achieve responsive performance, interactive graphics rely on lazy data
access and computing patterns, so the deferred mechanisms within
plyranges should help support interactive genomics applications.

@@

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages for
the code review they performed. This article was written with
\texttt{knitr} \cite{R-knitr} and the figures were made with
\texttt{ggbio} \cite{R-ggbio}. All code required to reproduce this
article is available at \url{https://github.com/sa-lee/plyranges-paper}.

>>>>

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

@@

\author{\Large Stuart Lee
\footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}\vspace{0.05in} \newline\normalsize\emph{}   \and \Large Dianne Cook \(^{\ref{monash}}\)\vspace{0.05in} \newline\normalsize\emph{}   \and \Large Michael Lawrence
\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}\vspace{0.05in} \newline\normalsize\emph{}  }

@@

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Stuart Lee
\footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia} () and Dianne Cook \(^{\ref{monash}}\) () and Michael Lawrence
\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} ()},
             pdfkeywords = {Bioconductor, Grammar, Genomes, Data Analysis},  
            pdftitle={plyranges: A grammar of genomic data transformation},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
\textbf{\authorfont Stuart Lee
\footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}} \hskip 15pt \emph{\small }   \par \textbf{\authorfont Dianne Cook \(^{\ref{monash}}\)} \hskip 15pt \emph{\small }   \par \textbf{\authorfont Michael Lawrence
\footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author}} \hskip 15pt \emph{\small }

@@

\vskip 8.5pt \noindent \emph{Keywords}: Bioconductor, Grammar, Genomes, Data Analysis \par

@@

\noindent  \hypertarget{background}{%
\section{Background}\label{background}}

@@

\hypertarget{genomic-relation-algebra}{%
\subsection{Genomic Relation Algebra}\label{genomic-relation-algebra}}

\hypertarget{data-model}{%
\subsubsection{Data Model}\label{data-model}}

@@

The \texttt{plyranges} DSL is built on the core Bioconductor data
structure GRanges, which is essentially a constrained table, with fixed
columns for the chromosome, start and end coordinates, and the strand,
along with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment (figure
\ref{fig:GRanges}). GRanges balances flexibility with formal
constraints, so that it is applicable to virtually any genomic workflow,
while also being semantically rich enough to support high-level
operations on genomic ranges. As a core data structure, GRanges enables
interoperability between \texttt{plyranges} and the rest of
Bioconductor. Adhering to a single data structure simplifies the API and
makes it easier to learn and understand, in part because operations
become endomorphic, i.e., they return the same type as their input.

GRanges follows the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable describing the
observations. GRanges specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
dplyr. Thus, we define our algebra as an extension of the \texttt{dplyr}
algebra, and borrow its syntax conventions and design principles.

@@

\hypertarget{algebraic-operations}{%
\subsubsection{Algebraic operations}\label{algebraic-operations}}

@@

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including \emph{intersect}, \emph{union},
\emph{difference}, and \emph{complement}. There are two sets of these:
parallel and merging. For example, the parallel intersection (\emph{x
\%intersect\% y}) finds the intersecting range between \emph{xi} and
\emph{yi} for \emph{i} in \emph{1\ldots{}n}, where \emph{n} is the
length of both \emph{x} and \emph{y}. In contrast, the merging
intersection (\emph{intersect\_ranges(x, y)}) returns a new set of
disjoint ranges representing wherever there was overlap between a range
in \emph{x} and a range in \emph{y}. Finding the parallel union will
fail when two ranges have a gap, so we introduce a \emph{span} operator
that takes the union while filling any gap. The \emph{complement}
operation is unique in that it is unary. It finds the regions not
covered by any of the ranges in a single set. Closely related is the
\emph{between} parallel operation, which finds the gap separating
\emph{xi} and \emph{yi}. The binary operations are callable from within
arithmetic, restriction and aggregation expressions.

@@

\hypertarget{developing-workflows-with-plyranges}{%
\subsection{\texorpdfstring{Developing workflows with
\texttt{plyranges}}{Developing workflows with plyranges}}\label{developing-workflows-with-plyranges}}

@@

\hypertarget{peak-finding}{%
\subsubsection{Peak Finding}\label{peak-finding}}

@@

\hypertarget{computing-windowed-statistics}{%
\subsubsection{Computing Windowed
Statistics}\label{computing-windowed-statistics}}

@@

\hypertarget{quality-control-metrics}{%
\subsubsection{Quality Control Metrics}\label{quality-control-metrics}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

We have aimed for \texttt{plyranges} to have a simple and direct mapping
to the user's cognitive model, i.e., how the user thinks about the data.
This requires careful selection of the level of abstraction so that the
user can express workflows in the language of genomics. This motivates
the adoption of the tidy GRanges object as our central data structure.
The basic data.frame and \texttt{dplyr} tibble lack any notion of
genomic ranges and so could not easily support our genomic grammar, with
its specific verbs for range-oriented data manipulation. Another example
of cognitive consistency is how \texttt{plyranges} is insensitive to
direction/strand by default when, e.g., detecting overlaps.
\texttt{GenomicRanges} has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

A function is cohesive if it performs a singular task without producing
any side-effects. Singular tasks can always be broken down further at
lower levels of abstraction. For example, to resize a range, the user
needs to specify which position (start, end, midpoint) should be
invariant over the transformation. The \texttt{resize()} function from
the \texttt{GenomicRanges} package has a \texttt{fix} argument that sets
the anchor, so calling \texttt{resize()} coalesces anchoring and width
modification. The coupling at the function call level is justified since
the effect of setting the width depends on the anchor. However,
\texttt{plyranges} increases cohesion and decouples the anchoring into
its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names as
the primary means of expression, instead of a more complex mixture of
function and argument names. This results in the user being able to
conceptualize the \texttt{plyranges} DSL as a flat catalog of functions,
without having to descend further into documentation to understand a
function's arguments. A flat function catalog also enhances API
discoverability, particularly through auto-completion in IDEs. One
downside of pushing cohesion to this extreme is that function calls
become coupled, and care is necessary to treat them as a group when
modifying code.

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect in
the original \texttt{GenomicRanges} API. In other cases, the
\texttt{plyranges} API increases verbosity for the sake of clarity and
cohesion. Explicitly calling \emph{anchor()} can require more typing,
but the code is easier to comprehend. Another example is the set of
routines for importing genomic annotations, including
\emph{read\_gff()}, \emph{read\_bed()}, and \emph{read\_bam()}. Compared
to the generic \emph{import()} in \texttt{rtracklayer}, the explicit
format-based naming in \texttt{plyranges} clarifies intent and the type
of data being returned. Similarly, every \texttt{plyranges} function
that computes with strand information indicates its intentions by
including suffixes such as \emph{directed}, \emph{upstream} or
\emph{downstream} in its name, otherwise strand is ignored. The
\texttt{GenomicRanges} API does not make this distinction explicit in
its function naming, instead relying on a parameter that defaults to
strand sensitivity, an arguably unintuitive behavior.

A caveat to constructing a compatible interface with \texttt{dplyr} is
that \texttt{plyranges} makes extensive use of non-standard evaluation
in R via the \texttt{rlang} package \cite{R-rlang}. Simply, this means
that computations are evaluated in the context of the GRanges objects.
Both \texttt{dplyr} and \texttt{plyranges} are based on the
\texttt{rlang} language, because it allows for more expressive code that
is free of repeated references to the container. Implicitly referencing
the container is particularly convenient when programming interactively.
Consequently, when programming with \texttt{plyranges}, a user needs to
generally understand the \texttt{rlang} language and how to adapt their
code accordingly. Users familiar with the tidyverse should already have
such knowledge.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

@@

As the \texttt{plyranges} interface encourages tidy data practices, it
integrates well with the grammar of graphics \cite{Wickham2016-gz}. To
achieve responsive performance, interactive graphics rely on lazy data
access and computing patterns, so the deferred mechanisms within
\texttt{plyranges} should help support interactive genomics
applications.

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{data-availability}{%
\subsection{Data Availability}\label{data-availability}}

The BigWig file for the H3K27Me3 primary T CD8+ memory cells from
peripheral blood ChIP-seq data was downloaded from the
\texttt{AnnotationHub} package (2.12.0) under accession AH33458. The BAM
file corresponding to the H1 cell line ChIP-seq data is available at GEO
under accession GSM433167. The SNP array data for the H1 cell line data
is available at GEO under accession GSM1463263.

\hypertarget{software-versions}{%
\subsection{Software Versions}\label{software-versions}}

To produce the workflows as described in results section we used R
version 3.5 with the development version of \texttt{plyranges} (1.1.4)
and the \texttt{BioStrings} package (2.48.0) installed.

All code required to reproduce this article is available at
\url{https://github.com/sa-lee/plyranges-paper}.

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages for
the code review they performed. This article was written with
\texttt{knitr} \cite{R-knitr} and the figures were made with
\texttt{ggbio} \cite{R-ggbio}.
